Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
deploy the new intro

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a high-level, dynamically-typed
``scripting'' or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

\emph{[Alternate introduction, trying to tie the points more closely
to the content of the paper:\\
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
The performance of our new implementation is comparable to the
performance of~C++ (\secref{perf}).
\item
As expected, higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
Haskell is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time (\secref{penumbra}).
We~found it difficult to tell which is which in advance.
\item
It is possible to use Haskell successfully for research even with a
very modest background in functional programming. We found our Haskell
implementation to be more flexible about adding new research ideas
than our past C++ code (\secref{comparo}).
\item
The only significant impediment that Haskell has presented to our
research has been the difficulty of understanding and tuning the
performance of Haskell programs.
\end{itemize}
]}

@@

\subsection{Performance}
\seclabel{perf}

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a high-level, dynamically-typed
``scripting'' or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on an attempt to replace both these families
of languages with Haskell: 
\begin{itemize}
\item
Although had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
The performance of our new implementation is comparable to the
performance of~C++, and parallel performance was easy to achieve
(\secref{perf}). 
\item
As expected, higher-order functions made it unusually easy to
implement and experiment with a family of related algorithms
(\secref{hofs}).
\item
Haskell is surrounded by a penumbra of libraries and tools that
promise powerful abstractions.
Some deliver very high leverage, but others are not ready for prime
time.
We~found it difficult to tell which is which in advance (\secref{penumbra}).
\item
It is possible to use Haskell successfully for research even with a
very modest background in functional programming. 
Modifying our group's C++ code to try out new research ideas has been
difficult;
trying new ideas in Haskell has been easy
(\secref{comparo}).
\item
The only significant impediment that Haskell has presented to our
research has been the difficulty of understanding and improving the
performance of Haskell programs (\secref{awkward-profiling}).
\end{itemize}

@@

\subsection{Performance}

\seclabel{perf}
